---
title: "CX2_Clustering"
author: "YYY"
date: "9/14/2022"
output: html_document
---

## R Markdown
This markdown is used to cluster cell features extracted from CellXpress2 (CX2).
Before commencing, install these packages if necessary:
install.packages(c("dplyr","tidyverse","matrixStats","RColorBrewer","reshape2","ggplot2","corrplot","uwot","viridis","ggpubr"))
BiocManager::install(c("flowCore","Biobase","FlowSOM"))
remotes::install_github("JonathanIrish/MEMv3")

Import these libraries. They are required for this markdown file.
```{r, echo=FALSE, warning=FALSE, error=FALSE}
# Retrieve packages
library(dplyr)
library(tidyverse)
library(matrixStats)
library(RColorBrewer)
library(reshape2)
library(ggplot2)
library(corrplot)
library(uwot)
library(viridis)
library(ggpubr)
library(Biobase)
library(FlowSOM)
library(MEM)
```

Import the extracted cell features .csv file.
```{r}
# Set file directory
directory = "/Users/user/Desktop/CellXpress/"
filename = "Tonsil_cX2_features.csv"
file = paste(directory,filename, sep="")
file

# Import .csv data file
cellFeatures = read.csv(file)
head(cellFeatures)
```

CX2 extracts many features that are not necessary for clustering, and they can be removed first.
```{r}
# Remove redundant columns from all the exported features
col_exclude = c("Slide","total_intensity","stddev","skewness","kurtosis","pct")
cellFeatures_trimmed = select(cellFeatures, -contains(col_exclude))

# Renaming "Region" to "Well" to follow the 96-well plate format in CX2
names(cellFeatures_trimmed)[names(cellFeatures_trimmed) == "Region"] <- "Well"

colnames(cellFeatures_trimmed)
```

Verify that for all FOVs, each cell has a nuclear signal.
```{r}
# Check all cells for nuclear signal; the value should not be zero
nuclear_min_signal = cellFeatures_trimmed %>%
  group_by(Well) %>% 
  summarise(Minimum_DNA=min(mean_intensity.DNA.dna_region)) # Rename the nuclear channel if ever necessary

nuclear_min_signal
```

On a separate .csv, organize the cell region of each marker to be used (cell_region / dna_region / nondna_region), thenT import the .csv names.
For benchmarking purposes, two .csv files files are made: one with signals from specified cell regions and another with signals from the whole cell.
```{r}
# Import marker names file
# markers_filename = "CX2_Markers.cell.csv"
markers_filename = "CX2_Markers.region.csv"
markers_file = paste(directory,markers_filename, sep="")
cellMarkers = read.csv(markers_file)

cellMarkers
```

In CX2, the "mean_intensity" columns represent signals that are already scaled to their respective cellular areas.
Therefore, we just need to extract the relevant features using our marker names .csv file.
```{r}
# Merge marker names to be consistent with CX2's naming system
marker_name = cellMarkers$Marker
marker_region = cellMarkers$Region
features_to_use = paste(marker_name,marker_region, sep=".") # Rename the nuclear channel if ever necessary
features_to_use

# Pull out relevant features here; the "Well" column is needed to group cells by their FOVs later
selected_features = select(cellFeatures_trimmed, contains(features_to_use))
Well = cellFeatures_trimmed$Well

selected_features = cbind(Well,selected_features)
head(selected_features)
```

For each FOV, normalize each marker (except nucleus) relative to the FOV's median nuclear signal.
```{r}
# Obtain median nuclear signal per FOV
nuclear_median_signal = cellFeatures_trimmed %>%
  group_by(Well) %>% 
  summarise(Median_DNA=median(mean_intensity.DNA.dna_region))
nuclear_median_signal

# Merge median nuclear signal into the selected features table
selected_features_median = merge(nuclear_median_signal, selected_features, by = "Well")
head(selected_features_median)

# Normalize each FOV to its median nuclear signal
cellFeatures_norm = selected_features_median[,3:ncol(selected_features_median)] / selected_features_median[,2]
head(cellFeatures_norm)

# Add back "DNA" columns
DNA = cellFeatures_trimmed$mean_intensity.DNA.cell_region
cellFeatures_norm = cbind(DNA, cellFeatures_norm)
head(cellFeatures_norm)
```

Rename all the features such that they are more readable.
```{r}
# Merge DNA column and rename all columns
cellFeatures_name = append("DNA",cellMarkers$Name)
colnames(cellFeatures_norm) = cellFeatures_name

head(cellFeatures_norm)
colnames(cellFeatures_norm)
```

Set values percentile for thresholding and perform arcsinh transformation.
```{r}
# Set percentile threshold (two-sided cutoff)
percentile = 99
percentileBot = 1-percentile/100
percentileTop = percentile/100

# arcsinh transformation: cofactor 150 for fluorescence microscopy, 5 for mass cytometry
cofactor = 150
cellFeatures_trans = asinh(cellFeatures_norm / cofactor)
head(cellFeatures_trans)
```

Perform percentile thresholding for arcsinh transformed data: threshold across all FOVs.
```{r}
# Untransformed data: percentile thresholding
## Find bot and top quantiles
FOV_norm_matrix = data.matrix(cellFeatures_norm)
range_norm = colQuantiles(FOV_norm_matrix, probs = c(percentileBot, percentileTop))
quantileBot_norm = range_norm[,1]
quantileTop_norm = range_norm[,2]
## Rescale data to [0,1] while considering percentile threshold
cellFeatures_norm_rescale = sweep(FOV_norm_matrix, 2, quantileBot_norm, '-')
cellFeatures_norm_rescale = sweep(cellFeatures_norm_rescale, 2, (quantileTop_norm-quantileBot_norm), '/')
cellFeatures_norm_rescale[cellFeatures_norm_rescale<0] = 0
cellFeatures_norm_rescale[cellFeatures_norm_rescale>1] = 1

# Transformed data: percentile thresholding
## Find bot and top quantiles
FOV_trans_matrix = data.matrix(cellFeatures_trans)
range_trans = colQuantiles(FOV_trans_matrix, probs = c(percentileBot, percentileTop))
quantileBot_trans = range_trans[,1]
quantileTop_trans = range_trans[,2]
## Rescale data to [0,1] while considering percentile threshold
cellFeatures_trans_rescale = sweep(FOV_trans_matrix, 2, quantileBot_trans, '-')
cellFeatures_trans_rescale = sweep(cellFeatures_trans_rescale, 2, (quantileTop_trans-quantileBot_trans), '/')
cellFeatures_trans_rescale[cellFeatures_trans_rescale<0] = 0
cellFeatures_trans_rescale[cellFeatures_trans_rescale>1] = 1

# Quick data check
cellFeatures_norm_rescale = data.frame(cellFeatures_norm_rescale)
cellFeatures_trans_rescale = data.frame(cellFeatures_trans_rescale)
head(cellFeatures_norm_rescale)
head(cellFeatures_trans_rescale)
```

Plot graphs for (i) no arcsinh trans, (ii) arcsinh trans, (iii) no arcsinh trans + pct threshold, (iv) arcsinh trans + pct threshold. This serves as a visual check of the data. Ideally, the signal intensities for all markers would be around the same range after arcsinh transformation, and a "tip" at x = 1 after percentile thresolding is indicative of positive signals for that marker.
```{r, fig.width=8, fig.height=8, warning=FALSE}
# Color palette for graphs
color_palette = brewer.pal(n=9, name='Set1')
colors = colorRampPalette(color_palette)

# Add back "Well" column
Trans_n = cbind(Well, cellFeatures_norm)
Trans_y = cbind(Well, cellFeatures_trans)
Trans_N = cbind(Well, cellFeatures_norm_rescale)
Trans_Y = cbind(Well, cellFeatures_trans_rescale)

# Change the data into a format compatible with ggplot
reshape_1 = melt(Trans_n,
                 id.vars = "Well",
                 variable.name = "marker",
                 value.name = "signal")
reshape_2 = melt(Trans_y,
                 id.vars = "Well",
                 variable.name = "marker",
                 value.name = "signal")
reshape_3 = melt(Trans_N,
                 id.vars = "Well",
                 variable.name = "marker",
                 value.name = "signal")
reshape_4 = melt(Trans_Y,
                 id.vars = "Well",
                 variable.name = "marker",
                 value.name = "signal")

# Plot graphs per FOV
FOV_num = length(unique(Well))

plot_1 = ggplot(reshape_1) +
  geom_density(aes(x = signal, color = Well, group = Well)) +
  facet_wrap(vars(marker), scales = "free", ncol=7) +
  theme_bw() +
  scale_color_manual(values=colors(FOV_num))
plot_2 = ggplot(reshape_2) +
  geom_density(aes(x = signal, color = Well, group = Well)) +
  facet_wrap(vars(marker), scales = "free", ncol=7) +
  theme_bw() +
  scale_color_manual(values=colors(FOV_num))
plot_3 = ggplot(reshape_3) +
  geom_density(aes(x = signal, color = Well, group = Well)) +
  facet_wrap(vars(marker), scales = "free", ncol=7) +
  theme_bw() +
  scale_color_manual(values=colors(FOV_num))
plot_4 = ggplot(reshape_4) +
  geom_density(aes(x = signal, color = Well, group = Well)) +
  facet_wrap(vars(marker), scales = "free", ncol=7) +
  theme_bw() +
  scale_color_manual(values=colors(FOV_num))

plot_1
plot_2
plot_3
plot_4
```

Remove non-working markers: check for poor staining & weird data transformations.
From henceforth, we will only use data that are (iv) arcsinh trans + pct threshold.
```{r}
# Make sure the names of the markers correspond to the style CX2 outputs
exclude_features = c("T.bet","TCR.g.d","CD16","EGFR","CCR6","MCT","MMP.9","CD2")

# Remove features from data
drop_index = match(exclude_features, colnames(Trans_Y))
cellFeatures_purified = Trans_Y[, -c(drop_index)]

head(cellFeatures_purified)
```

Select markers for clustering Do not select all markers; only use those that are useful for phenotyping.
```{r}
# Make sure that the marker name spellings are consistent with the column names
UMAP_features = c("CK",
                  #"Vimentin",
                  "Podoplanin",
                  "CD31",
                  "a.SMA",
                  "Collagen.IV",
                  "CD11c",
                  "CD11b",
                  "CD68",
                  "CD163",
                  "CD206",
                  #"CD2",
                  #"CD5",
                  "CD7",
                  "CD4",
                  "CD8",
                  "CD20",
                  "CD15",
                  "FOXP3"
                  )
#UMAP_features = c("CK","CD31","CD11b","CD20","CD5","CD7","CD4","CD8","FOXP3","CD68","CD163","CD206","CD11c","CD15","Podoplanin")

# Pull out the markers of interest
cellFeatures_UMAP = cellFeatures_purified[, c(UMAP_features)]

head(cellFeatures_UMAP)
```

OPTIONAL: Perform a correlation matrix on the selected UMAP markers for a brief visual check of the data.
```{r, fig.height=2, fig.width=6}
# Correlation across UMAP markers, and across all FOVs
corr_UMAP_features_all = cor(cellFeatures_UMAP, method = "spearman") 
corr_UMAP_features_all_plot = corrplot(corr_UMAP_features_all, tl.col = "black", order = "hclust", hclust.method = "average", tl.cex = 1)
corr_UMAP_features_all_plot
```

Run UMAP dimension reduction on the selected markers, then extract UMAP values. A 2D dimension reduction is sufficient for our purposes.
```{r}
set.seed(1) # Set seed to ensure consistent results after every run
UMAP_run = umap(cellFeatures_UMAP,
                n_neighbors = 15,
                n_components = 2,
                min_dist = 0.01,
                metric = 'euclidean',
                ret_model = TRUE, 
                verbose = TRUE)

# Pull out the UMAP values
UMAP_values = data.frame(UMAP_run$embedding)
colnames(UMAP_values) = c("UMAP1", "UMAP2")
head(UMAP_values)

# Difference between the maximum and minimum values of UMAP1 and UMAP2
UMAP_min_max = apply(UMAP_values,2, range)
UMAP_range = apply(UMAP_min_max,2, diff)
UMAP_range
```

Generate a density plot and dot plot for the UMAP values.
```{r, fig.height=4, fig.width=8}
# Scaling the aspect ratio for the graph
UMAP_plot_axis_ratio = max(UMAP_range) / min(UMAP_range)

# Plot without coloring by FOV
UMAP_Plot = ggplot(UMAP_values) +
  geom_point(aes(x=UMAP1, y=UMAP2), cex = 0.5) +
  coord_fixed(ratio = UMAP_plot_axis_ratio) +
  labs(x = "UMAP 1", y = "UMAP 2", title = "UMAP Plot") +
  theme_bw() +
  theme(axis.text=element_blank(),
        axis.ticks=element_blank()
        )

# Plot with coloring by FOV
FOV_num = length(unique(Well))

UMAP_values_FOV = cbind(UMAP_values, Well) # Need to add back "Well" column to define FOVs
UMAP_Plot_FOV = ggplot(UMAP_values_FOV) +
  geom_point(aes(x=UMAP1, y=UMAP2, color=Well), cex = 0.5) +
  coord_fixed(ratio = UMAP_plot_axis_ratio) +
  labs(x = "UMAP 1", y = "UMAP 2", title = "UMAP Plot by FOV") +
  scale_color_manual(values=colors(FOV_num)) + # color palette for graphs is already defined previously
  theme_bw() +
  theme(axis.text=element_blank(),
        axis.ticks=element_blank()
        )

UMAP_Plot
UMAP_Plot_FOV
```

Perform FlowSOM clustering on the actual selected markers used for UMAP dimension reduction.
```{r, fig.height=8, fig.width=8}
# Enter the target number of clusters; over-clustering relative to the number of markers is a good idea as it helps downstream annotation
target.clusters = 10

# Make a new object, flowFrame, specifically for use in FlowSOM
## Step 1: set flowFrame exprs: this must be a matrix containing the dataset
flowFrame_exprs = as.matrix(cellFeatures_UMAP)
#head(flowFrame_exprs)

## Step 2: set flowFrame parameters: this must be an 'AnnotatedDataFrame' containing information about each column of the flowFrame_exprs data
### The information required are: name, desc, range, minRange, maxRange
### The information should also match the colnames of the flowFrame_exprs matrix
flowFrame_parameters = data.frame(name = colnames(flowFrame_exprs),
                                  desc = colnames(flowFrame_exprs),
                                  range = apply( apply(flowFrame_exprs, 2, range) , 2, diff),
                                  minRange = apply(flowFrame_exprs, 2, min),
                                  maxRange = apply(flowFrame_exprs, 2, max)
                                  )
#head(flowFrame_parameters)

## Step 3: Make the flowFrame object
flowFrame = new("flowFrame",
                exprs = flowFrame_exprs,
                parameters = AnnotatedDataFrame(flowFrame_parameters)) 
#head(flowFrame)

# Implement FlowSOM on the flowFrame
fsom = FlowSOM(flowFrame,
               colsToUse = 1:ncol(flowFrame),
               nClus = target.clusters,
               seed = 1)
#head(fsom)

# This gets the cluster information for each cell
FlowSOM_clusters = GetMetaclusters(fsom)
#head(fsom_clusters)
```

Plot FlowSOM clusters onto the UMAP plots.
```{r, fig.height=6, fig.width=12}
# Merge FlowSOM clusters with UMAP values
FlowSOM_UMAP = cbind(FlowSOM_clusters, UMAP_values)

# 1. Plot FlowSOM clusters without considering FOVs
Clusters_num = target.clusters

FlowSOM_UMAP_Plot = ggplot(FlowSOM_UMAP) +
  geom_point(aes(x=UMAP1, y=UMAP2, color=FlowSOM_clusters), cex = 0.5) +
  coord_fixed(ratio = UMAP_plot_axis_ratio) +
  labs(x = "UMAP 1", y = "UMAP 2", color="Clusters", title = "FlowSOM Clusters on UMAP") +
  scale_color_manual(values=colors(Clusters_num)) + # color palette for graphs is already defined previously
  theme_bw() +
  theme(axis.text=element_blank(),
        axis.ticks=element_blank()
        )



# 2. Plot FlowSOM clusters, this time while considering FOVs
## Add back "Well" column, then pull out well names
FlowSOM_UMAP_FOV = cbind(Well, FlowSOM_UMAP)
FOV_names = unique(Well)

## Subset data based on FOV and store each dataframe in a list
FlowSOM_UMAP_FOV_subsetted = list()
for (u in 1:length(FOV_names)){
  FlowSOM_UMAP_FOV_subset = subset(FlowSOM_UMAP_FOV, Well == FOV_names[u])
  FlowSOM_UMAP_FOV_subset = FlowSOM_UMAP_FOV_subset[, -c(1)] # Remove "Well" column as it is a string and cannot be plotted
  FlowSOM_UMAP_FOV_subsetted[[u]] = FlowSOM_UMAP_FOV_subset
}

## Plot FlowSOM clusters as before, for each dataframe within the list
FSOM_UMAP_FOV_plt = list()
for (v in 1:length(FOV_names)){
  ### Extract each FOV's dataframe from the list
  FlowSOM_UMAP_FOV_subsetted_df = as.data.frame(FlowSOM_UMAP_FOV_subsetted[v])
  
  ### Plot FlowSOM cluster for each FOV's dataframe
  FlowSOM_UMAP_FOV_df_Plot = ggplot(FlowSOM_UMAP_FOV_subsetted_df) +
    geom_point(aes(x=UMAP1, y=UMAP2, color=FlowSOM_clusters), cex = 0.5) +
    coord_fixed(ratio = UMAP_plot_axis_ratio) +
    labs(x = "UMAP 1", y = "UMAP 2", color="Clusters", title = "FlowSOM Clusters on UMAP by FOV") +
    scale_color_manual(values=colors(Clusters_num)) + # color palette for graphs is already defined previously
    theme_bw() +
    theme(axis.text=element_blank(),
          axis.ticks=element_blank()
          )
  
  ### Save the FlowSOM plots to the empty list
  FSOM_UMAP_FOV_plt[[v]] = FlowSOM_UMAP_FOV_df_Plot
}



# 3. Generate FlowSOM plots
## To facilitate comparison, here we append the pan-FOV FlowSOM plot to the list of per-FOV FlowSOM plot
FlowSOM_UMAP_FOV_plot_all = append(list(FlowSOM_UMAP_Plot), FSOM_UMAP_FOV_plt)

## Pan-FOV FlowSOM plot
FlowSOM_UMAP_Plot

## Per-FOV FLowSOM plot: in each graph, the first plot is pan-FOV and the other plots are in chronological FOV order
FlowSOM_UMAP_FOV_plot = ggarrange(plotlist = FlowSOM_UMAP_FOV_plot_all, ncol=3)
FlowSOM_UMAP_FOV_plot
```

Make MEM plots (Marker Enrichment Modeling) based on the cell clusters --- across all FOVs.
Here, the relative marker expression of each cell cluster is extracted, thereby enabling cell cluster annotation.
```{r, fig.height=5, fig.width=4}
# Merge cluster numbers with the markers selected prior to UMAP
## Do not change the syntax below as MEM needs to use this exact format
cluster = as.numeric(FlowSOM_clusters)
MEM_data = cbind(cellFeatures_UMAP, cluster)
head(MEM_data)

# Run MEM
## Scale zero = off (zero.ref = FALSE)
MEM.data_zeroOFF = MEM(
  MEM_data,
  transform = FALSE,
  cofactor = 1, # this only matters if transform is on
  choose.markers = FALSE,
  markers = "all",
  choose.ref = FALSE,
  zero.ref = FALSE, # Scale zero = on
  rename.markers = FALSE,
  new.marker.names = "none",
  file.is.clust = FALSE,
  add.fileID = FALSE,
  IQR.thresh = NULL
  )
## Scale zero = on (zero.ref = TRUE)
MEM.data_zeroON = MEM(
  MEM_data,
  transform = FALSE,
  cofactor = 1, # this only matters if transform is on
  choose.markers = FALSE,
  markers = "all",
  choose.ref = FALSE,
  zero.ref = TRUE, # Scale zero = on
  rename.markers = FALSE,
  new.marker.names = "none",
  file.is.clust = FALSE,
  add.fileID = FALSE,
  IQR.thresh = NULL
  )

# Build MEM heatmaps
## Heatmap for scale zero = off
build.heatmaps(
  MEM.data_zeroOFF,
  cluster.MEM = "both",
  cluster.medians = "none",
  cluster.IQRs = "none",
  display.thresh = 1,
  newWindow.heatmaps = F,
  output.files = F, # it will still create an empty "output files" folder
  labels = F, # removes the annoying labels an the right side of the figure
  only.MEMheatmap = T
  )
## Heatmap for scale zero = on
build.heatmaps(
  MEM.data_zeroON,
  cluster.MEM = "both",
  cluster.medians = "none",
  cluster.IQRs = "none",
  display.thresh = 1,
  newWindow.heatmaps = F,
  output.files = F, # it will still create an empty "output files" folder
  labels = F, # F removes the annoying labels an the right side of the figure
  only.MEMheatmap = T
  )
```

Export clustered values into a clusters.csv file.
A template for manual annotation is provided in the next section (be very careful when running it).
```{r}
# Save file without separating cells by FOV
## Merge cluster value to each cell
cellID = data.frame(cellID = cellFeatures_trimmed$label.mask.cell_region) # Rename cell region if ever necessary
clusterID = data.frame(cluster_ID = cluster)
cell_cluster = cbind(Well, cellID, cellFeatures_UMAP, clusterID)
head(cell_cluster)

## Set filename and save directory
cluster_filename = "CX2_TonsilClusters.csv"
cluster_output = paste(directory,cluster_filename, sep="")

## Write cluster.csv
write.csv(cell_cluster, cluster_output)
```

Color each cell cluster to faciliate manual annotation.
```{r, warning=FALSE, fig.height=3, fig.width=4}
# Extract FOV, cluster ID, x coordinate, y coordinate, cell area, and cell ID
cell_mask_features = data.frame(Well = cellFeatures_trimmed$Well,
                                cellID,
                                cluster_ID = cluster,
                                pos_x = cellFeatures_trimmed$pos_x.mask.cell_region,
                                pos_y = cellFeatures_trimmed$pos_y.mask.cell_region,
                                cell_Area = cellFeatures_trimmed$area.mask.cell_region
                                )

# Subset cells based on FOVs
FOV_names = unique(Well)
cell_mask_subsetted = list()
for (m in 1:length(FOV_names)){
  ## Subsetting cells per FOV
  cell_mask_subset = subset(cell_mask_features, Well == FOV_names[m])
  cell_mask_subset = cell_mask_subset[, -c(1)] # Remove "Well" column as it is a string and cannot be plotted
  cell_mask_subsetted[[m]] = cell_mask_subset
}

# Make cell masks cells based on FOVs, then save in custom folder
for (n in 1:length(FOV_names)){
  ## Make new folder to save per-FOV plots
  ### Set custom filename
  cluster_FOV_name = FOV_names[n]
  cell_mask_FOV_folder = paste("CX2_TonsilClusters_",cluster_FOV_name,"_mask/", sep="")
  ### Set new directory
  directory_new = paste(directory,cell_mask_FOV_folder, sep="")
  ### Set new path using file.path
  path_new = file.path(dirname(directory_new), cell_mask_FOV_folder)
  ### Make new folder
  dir.create(path_new)
  
  ## Extract each FOV's dataframe from the list
  cell_mask_df = as.data.frame(cell_mask_subsetted[n])
  
  ## Make cell mask plot of the FOV
  cell_mask_plot = ggplot(cell_mask_df) +
    geom_point(aes(x=pos_x, y=pos_y, size=cell_Area),
               color='mediumpurple2',
               show.legend=FALSE
               ) +
    scale_size_continuous(range = c(0.1, 3.6)) +
    coord_fixed(ratio = 1) +
    scale_y_reverse() + #need to reverse the y=axis as R seems to read the y axis opposite
    theme_bw() +
    theme(axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank()
          )
  
  ## Now, add the cell color for each cluster, and then write out as individual files
  for (o in 1:target.clusters){
    ### Subsetting FOV dataframe by cell cluster
    cell_mask_df_cluster = subset(cell_mask_df, cluster_ID == o)
    
    ### Add cell mask plot with per-cluster coloring
    cell_mask_plot_cluster = cell_mask_plot +
      geom_point(data = cell_mask_df_cluster,
                 aes(x=pos_x, y=pos_y, size=cell_Area),
                 color='chocolate',
                 show.legend=FALSE
                 ) +
      scale_size_continuous(range = c(0.1, 3.6)) +
      coord_fixed(ratio = 1) +
      scale_y_reverse() + #need to reverse the y=axis as R seems to read the y axis opposite
      theme_bw() +
      theme(axis.title=element_blank(),
            axis.text=element_blank(),
            axis.ticks=element_blank()
            )
    
    ### Save into new directory
    #### Graph name and filetype
    saveplot_cluster = o
    saveplot_name = paste("Cluster_",saveplot_cluster,".png", sep="") #save as png
    
    ggsave(saveplot_name,
           path = path_new)
  }
}

# Save the cell_mask_features file; it is very useful for future mapping
## Set filename and save directory
cell_mask_features_filename = "CX2_TonsilCellMaskFeatures.csv"
cell_mask_features_output = paste(directory,cell_mask_features_filename, sep="")

write.csv(cell_mask_features, cell_mask_features_output)
head(cell_mask_features)
```

Here is a custom template .csv for manual annotation. BE CAREFUL WHEN RUNNING THIS TO NOT OVERWRITE ANNOTATIONS THAT YOU HAVE BEEN WORKING ON! 
Change file name, directory, or make backups accordingly if necessary.
```{r}
# PAUSE POINT: read the section before proceeding!
PAUSE_POINT
```

```{r}
# Make empty data.frame for manual annotation
blank_vector = rep('',target.clusters)
annotation_cluster = data.frame(cluster_ID = c(1:target.clusters),
                                celltype = c(blank_vector),
                                comments = c(blank_vector),
                                celltypes_currently_used = c(blank_vector)
                                )

# Set filename and save directory
annotation_template = "CX2_TonsilAnnotation.csv"
annotation_output = paste(directory,annotation_template, sep="")

# Write template .csv
write.csv(annotation_cluster, annotation_output)
```

This error message is placed here on purpose as a stop point.
Return to this script only after the cells are fully annotated.
```{r}
STOP_POINT
```

After annotating all cells, it is time to recolor them to produce a phenotype map.
First, merge the annotated clusters to the cell mask features.
```{r}
# Import the annotation and cell mask .csv files. Check the file names to ensure that they are correct.
filename.annotated = "CX2_TonsilAnnotation.csv"
file.annotated = paste(directory,filename.annotated, sep="")
file.annotated

filename.cellmask = "CX2_TonsilCellMaskFeatures.csv"
file.cellmask = paste(directory,filename.cellmask, sep="")
file.cellmask

# Read the files, then remove the extra column that appears during importing
file.a = read.csv(file.annotated)
file.cm = read.csv(file.cellmask)

file.a = file.a[, -1]
file.cm = file.cm[, -1]

head(file.a)
head(file.cm)

# Merge the files by cluster_ID
file_for.plotting = merge(file.a,file.cm, by="cluster_ID")
head(file_for.plotting)

##Ensure that every cell is annotated; both output values should be the same
file_for.plotting %>%
  summarise(Number.cells=length(cellID), Number.annotated=length(celltype))
```

Plot the cell phenotype map. Cell masks, with each cell type being colored, will also be produced.
```{r, warning=FALSE, fig.height=2, fig.width=4}
# Get color palette
mapcolor_palette = brewer.pal(n=12, name='Paired')

# Subset the FOVs
FOV_names = unique(Well)
file.plot_subsetted = list()

for (f in 1:length(FOV_names)){
  ## Subsetting cells per FOV
  file.plot_subset = subset(file_for.plotting, Well == FOV_names[f])
  file.plot_subsetted[[f]] = file.plot_subset
}

# Plot celltype-specific masks and a full phenotype map
celltypes = unique(file.plot_subset$celltype)
celltypes_num = length(celltypes)

phenotype.map_list = list()

for (g in 1:length(FOV_names)){
  ## Make new folder to save per-FOV plots
  ### Set custom filename
  map_FOV_name = FOV_names[g]
  phenotype_map_folder = paste("CX2_Tonsil_Phenotype.Map_",map_FOV_name, sep="")
  ### Set new directory
  directory_map = paste(directory,phenotype_map_folder, sep="")
  ### Set new path using file.path
  path_map = file.path(dirname(directory_map), phenotype_map_folder)
  ### Make new folder
  dir.create(path_map)
  
  ## Extract each FOV's dataframe from the list
  map_FOV_df = as.data.frame(file.plot_subsetted[g])
  
  ## Make a complete phenotype map
  phenotype_map = ggplot(map_FOV_df) +
    geom_point(aes(x=pos_x, y=pos_y, size=cell_Area, color=celltype),
               show.legend=TRUE) +
    guides(size = "none") +
    scale_size_continuous(range = c(0.1, 2.5)) +
    coord_fixed(ratio = 1) +
    scale_color_manual(values=mapcolor_palette) + #predefined colors above
    scale_y_reverse() + #need to reverse the y=axis as R seems to read the y axis opposite
    labs(title = "Cell Phenotype Map", color = "Cell Types") + #add title for the cell type and name the color coolumn accordingly
    theme_bw() +
    theme(axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank()
          )
  
  ## Save phenotype map into the new folder
  phenotype_map_name = paste("Cell Phenotype Map",".png", sep="") #save as png
  
  ggsave(phenotype_map_name,
         path = path_map)
  
  ## Store the phenotype maps into the phenotype.map_list for sanity check later
  phenotype.map_list[[g]] = phenotype_map
  
  ## Make celltype-specific masks, then save in custom folder
  for (h in 1:length(celltypes)){
    ### Get the current loop's cell name
    celltype_name = celltypes[h]
    
    ### Make a cell mask for the FOV
    celltype_plot = ggplot(map_FOV_df) +
      geom_point(aes(x=pos_x, y=pos_y, size=cell_Area),
                 color='mediumpurple2',
                 show.legend=FALSE
                 ) +
      scale_size_continuous(range = c(0.1, 3.6)) +
      coord_fixed(ratio = 1) +
      scale_y_reverse() + #need to reverse the y=axis as R seems to read the y axis opposite
      theme_bw() +
      theme(axis.title=element_blank(),
            axis.text=element_blank(),
            axis.ticks=element_blank()
            )
    
    ### Subset the dataframe such that only the relevant cell type is selected
    map_FOV_df_subset = subset(map_FOV_df, celltype == celltype_name)
    
    ### Add the color for each cell type
    celltype_coloring = celltype_plot +
      geom_point(data = map_FOV_df_subset,
                 aes(x=pos_x, y=pos_y, size=cell_Area, color=celltype),
                 color='chocolate',
                 show.legend=FALSE
                 ) +
      scale_size_continuous(range = c(0.1, 3.6)) +
      coord_fixed(ratio = 1) +
      scale_y_reverse() + #need to reverse the y=axis as R seems to read the y axis opposite
      labs(title = celltype_name) + #add title for the cell type
      theme_bw() +
      theme(axis.title=element_blank(),
            axis.text=element_blank(),
            axis.ticks=element_blank()
            )
    
    ### Save cell type color maps into the new folder
    celltype_plot_type = "png" # save as png
    celltype_plot_name = paste("Phenotype.Map_",celltype_name,".png", sep="")
    
    ggsave(celltype_plot_name,
          path = path_map)
  }
}

# Visual check for phenotype maps
phenotype.map_view = ggarrange(plotlist = phenotype.map_list)
phenotype.map_view
```


